---
title: "Karls_data"
author: "Kailin"
date: "2026-03-12"
output: html_document
---

```{r clean, include=FALSE}
rm(list = ls())
gc()
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
pacman::p_load(tidyverse, readxl, here, skimr, inspectdf, naniar, corrplot, tableone, patchwork, scales, broom, ResourceSelection, car)
```

```{r pptsetup, include=FALSE}
# Shared themes & colours (PPT optimisation: large font, white background, no unnecessary grid lines)
theme_ppt <- theme_bw(base_size = 16) +
  theme(
    plot.title       = element_text(face = "bold", size = 18),
    plot.subtitle    = element_text(size = 13, colour = "grey40"),
    legend.position  = "bottom",
    legend.title     = element_text(size = 13),
    legend.text      = element_text(size = 12),
    axis.title       = element_text(size = 14),
    axis.text        = element_text(size = 13),
    panel.grid.minor = element_blank(),
    strip.text       = element_text(face = "bold", size = 13)
  )

trt_colours <- c(A = "#0D7680", B = "#E8A838", C = "#A06030")

# Function to save images (to ../figs/, in 16:9 PowerPoint aspect ratio)
save_ppt_fig <- function(filename, width = 10, height = 6) {
  ggsave(
    filename = here("figs", filename),
    width = width, height = height, dpi = 200, bg = "white"
  )
}
```

# The Study

The study investigates the effect of three treatments (A, B, C) on AF and PR. Potential covariates include participant age, sex, smoking status, and individual SNP genotypes encoded under an additive genetic model.

# Read in Data

The dataset (`FITFR-final-data-v3.xlsx`) contains 100 participants and 58 variables, organised into three broad categories: demographics (`age`, `sex`, `treatment`, `smoker`), genetics (51 SNP variables with the prefix `rs`), and outcome measures (`AF` and `PR`).

## Load raw data

The raw file has a two-row header: the first row contains broad category labels ("demographics", "genetics", "measures") and the second row contains the actual column names. We skip the first row on read and drop the resulting category-label row and the empty leading column manually.

```{r load}
df_raw <- read_excel(here("raw-data", "FITFR-final-data-v3.xlsx"), sheet = 1, skip = 1)

cat("Raw dimensions:", nrow(df_raw), "x", ncol(df_raw))
```

# Cleaning / Pre-processing

## · Remove junk rows

The last four rows of the file are artefacts: three fully-empty rows and one garbled header fragment. Every genuine participant has a non-missing ID, so filtering on `ID` removes them cleanly.

```{r remove-junk}
df <- df_raw |>
  filter(!is.na(ID))

cat("Dimensions after removing junk rows:", nrow(df), "x", ncol(df))
```

## · Check column types

Before proceeding with cleaning, we inspect the column types to verify that variables have been read in correctly and identify any that require conversion.

```{r inspectdf_before}
# Variable Types Overview
inspect_types(df) |> show_plot()

# Frequency distributions of the categorical variables
inspect_cat(df)

# Distributions of the continuous variables
df |> select(-ID) |> inspect_num() |> show_plot()
```

The majority of columns (56) are read in as character, with only 2 numeric columns (`ID` and `AF`). This confirms that type conversion is required: `age` must be coerced to numeric, and categorical variables such as `sex`, `treatment`, `smoker`, and `PR` must be converted to factors. The SNP columns will also require some encoding under an additive genetic model.

## · Fix column types

Based on the type inspection above, `ID`, `age`, and `AF` are converted to their correct numeric types. Categorical variables are addressed in the following step.

```{r col-types}
df <- df |>
  mutate(
    ID  = as.integer(ID),
    age = as.numeric(age),
    AF  = as.numeric(AF)
  )
df
```

## · Standardise categorical variables

Several variables have inconsistent capitalisation or coding. We standardise them here and convert each to a factor.

```{r standardise-cats}
df <- df |>
  mutate(
    # sex: Female/female/Male/male/M → Female / Male
    sex = case_when(
      str_to_lower(sex) == "female"         ~ "Female",
      str_to_lower(sex) %in% c("male", "m") ~ "Male",
      TRUE                                  ~ NA_character_
    ),
    sex = factor(sex, levels = c("Female", "Male")),

    # treatment: a/b/c → A/B/C
    treatment = factor(str_to_upper(treatment), levels = c("A", "B", "C")),

    # smoker: only 'N' is present in this dataset
    smoker = factor(smoker, levels = "N"),

    # PR: yes/no → ordered factor
    PR = case_when(
      str_to_lower(PR) == "yes" ~ 1L,
      str_to_lower(PR) == "no"  ~ 0L,
      TRUE                      ~ NA_integer_
    ),
    PR = factor(PR, levels = c(0, 1), labels = c("no", "yes"))
  )
df
```

## · Sentinel value −99 in age

Two participants have `age = -99`, which is a sentinel for missing data. These are recoded to `NA`.

```{r age-sentinel}

df$age == -99

n_sentinel <- sum(df$age < 0, na.rm = TRUE)
cat(n_sentinel, "rows with age = -99 recoded to NA")

df <- df |>
  mutate(age = if_else(age == -99, NA_real_, age))
df
```

## · Zero-variance columns

Two columns carry no information for modelling:

-   `smoker`: all participants are non-smokers (`N`); no variation to estimate an effect from.
-   `rs123456`: all participants are `CC`; the SNP is monomorphic in this sample.

Both are retained in the cleaned dataset for transparency but should be excluded from any model.

```{r zero-var}
zero_var_cols <- c("smoker", "rs123456")
cat("Zero-variance columns (retain but exclude from models):",
    paste(zero_var_cols, collapse = ", "))
```

## · SNP additive encoding

SNP variables are encoded using a standard additive genetic model. Each genotype is converted into a numeric value representing the number of alternate alleles present.

The encoding follows these rules:

-   **0**: homozygous reference genotype (no alternate alleles)
-   **1**: heterozygous genotype (one alternate allele)
-   **2**: homozygous alternate genotype (two alternate alleles)

Because the raw data store genotypes as character pairs (e.g. `AA`, `AC`, `CC`), the encoding is performed separately for each SNP column. The procedure is:

-   Identify all **unique genotypes** observed for the SNP.
-   Extract **homozygous genotypes** (e.g. `AA`, `CC`, `GG`, `TT`).
-   Determine the **reference allele** as the allele appearing in the **most frequent homozygous genotype**.
-   Treat all other observed alleles at that SNP as **alternate alleles**.
-   For each genotype, **count the number of alternate alleles** it contains:
    -   e.g. if `A` is the reference allele
        -   `AA → 0`
        -   `AC → 1`
        -   `CC → 2`
-   Missing values remain as `NA`.

Finally, the encoded SNP values are stored as **ordered factors with levels `0 < 1 < 2`**, preserving the ordering of increasing alternate-allele dosage.

```{r snp-encode}
snp_cols <- names(df)[str_detect(names(df), "^rs")]

encode_snp <- function(x) {
  x       <- as.character(x)
  genos   <- na.omit(unique(x))
  homozyg <- genos[substr(genos, 1, 1) == substr(genos, 2, 2)]
  if (length(homozyg) < 1) return(factor(x))
  counts      <- table(x[x %in% homozyg])
  ref_allele  <- substr(names(which.max(counts)), 1, 1)
  alt_alleles <- setdiff(unique(unlist(strsplit(genos, ""))), ref_allele)
  count_alt   <- function(g) sum(strsplit(g, "")[[1]] %in% alt_alleles)
  coded       <- sapply(x, function(g) if (is.na(g)) NA_integer_ else count_alt(g))
  factor(coded, levels = c(0, 1, 2), ordered = TRUE)
}

df[snp_cols] <- lapply(df[snp_cols], encode_snp)
```

## · Cleaned dataset summary

Following all cleaning and encoding steps, we re-inspect the dataset to verify that all column types are now correct and that the cleaning steps have been applied as expected.

```{r inspectdf_after}
# Variable Types Overview
inspect_types(df) |> show_plot()

# Frequency distributions of the categorical variables
inspect_cat(df)

# Distributions of the continuous variables
df |> select(-ID) |> inspect_num() |> show_plot()
```

The types plot now shows the expected structure: 55 factor variables (including the SNP columns as ordered factors) and 3 numeric variables (`ID`, `age`, `AF`), confirming that all conversions were applied correctly.

Basic Statistical Summary:

```{r summary}
key_vars <- c("ID", "age", "sex", "treatment", "AF", "PR")

cat("Dimensions:", nrow(df), "x", ncol(df), "\n\n")

cat("Missing values per key variable:\n")
print(colSums(is.na(df[key_vars])))

cat("\nTreatment distribution:\n")
print(table(df$treatment, useNA = "ifany"))

cat("\nSex distribution:\n")
print(table(df$sex, useNA = "ifany"))

cat("\nPR distribution:\n")
print(table(df$PR, useNA = "ifany"))

cat("\nAF summary:\n")
print(summary(df$AF))

cat("\nAge summary:\n")
print(summary(df$age))
```

The cleaned dataset contains 100 participants across 58 variables. Only `age` has missing values (n = 2). Treatment groups are reasonably balanced (A: 32, B: 29, C: 39), as is sex (Female: 51, Male: 49) and PR outcome (no: 49, yes: 51). AF ranges from 26.02 to 95.99 with a mean of 61.32 (median: 61.06), age ranges from 25 to 45 with a mean of 34.16.

## · Save cleaned data

```{r save}
write.csv(df, here("data", "FITFR_clean.csv"), row.names = FALSE, na = "")
cat("Saved to", here("data", "FITFR_clean.csv"))
```

# EDA

[EDA]{.underline} = **Exploratory Data Analysis**

This involves ‘getting to know’ the data before formal modelling begins—examining **what the data looks like**, **whether there are any missing values**, and **what relationships exist between variables**.

## · Missing Values

Missing data were assessed across all variables to determine whether any imputation or case exclusion would be required prior to modelling.

```{r missing, fig.height=8}
gg_miss_var(df) +
  labs(title = "Missing values per variable") +
  theme_bw()
```

Only `age` contains missing values, with 2 observations recorded as sentinel value −99 and subsequently recoded to `NA` during cleaning. All other variables are complete, so no imputation is necessary.

## · Overview Summary

To obtain an overview of the dataset's structure and the distribution of key variables, we apply `skim()` across the cleaned data frame.

```{r skimr}
skim(df)
```

```{r descriptive-stats}
key_vars <- c("age", "sex", "treatment", "AF", "PR")

df |>
  select(all_of(key_vars)) |>
  summarise(across(
    where(is.numeric),
    list(
      mean = ~ round(mean(.x, na.rm = TRUE), 2),
      sd   = ~ round(sd(.x, na.rm = TRUE), 2),
      min  = ~ round(min(.x, na.rm = TRUE), 2),
      max  = ~ round(max(.x, na.rm = TRUE), 2),
      missing = ~ sum(is.na(.x))
    )
  ))

# Categorical variables
df |>
  select(all_of(key_vars)) |>
  select(where(is.factor)) |>
  pivot_longer(everything(), names_to = "variable", values_to = "value") |>
  group_by(variable, value) |>
  summarise(n = n(), .groups = "drop") |>
  mutate(percent = round(n / sum(n) * 100, 1), .by = variable)
```

The dataset contains 100 participants and 58 columns, comprising 55 factor variables and 3 numeric variables. Age is the only variable with missing values (n = 2, complete rate = 0.98), with a mean of 34.2 years (SD = 6.2). AF has no missing values, with a mean of 61.3 (SD = 19.8). Among the categorical variables, sex is approximately balanced (Female: 51, Male: 49), and treatment groups are reasonably well-distributed (C: 39, A: 32, B: 29). All participants are non-smokers, confirming that smoking status carries no variation and will be excluded from modelling. The SNP variables are fully observed and encoded as ordered factors with three levels (0, 1, 2), with heterozygous genotypes (dosage = 1) being the most common across most loci.

## · Distribution of variables

To understand the structure of the data, we examine the distribution of all variables, separating continuous variables from categorical ones.

### - Continuous Variables

The dataset contains two continuous variables of interest: `AF` (the primary outcome) and `age` (a potential covariate).

```{r continuous-dist}
# AF Distribution
p1 <- ggplot(df, aes(x = AF)) +
  geom_histogram(bins = 30, fill = "steelblue", alpha = 0.7) +
  geom_density(aes(y = after_stat(count)), colour = "red") +
  labs(title = "Distribution of AF") +
  theme_bw()

# Age Distribution
p2 <- ggplot(df, aes(x = age)) +
  geom_histogram(bins = 20, fill = "darkorange", alpha = 0.7) +
  labs(title = "Distribution of Age") +
  theme_bw()

p1 + p2
```

According to the distribution plots above, AF is roughly uniformly distributed across its range of approximately 26 to 96, with no strong skew or obvious outliers. Age is similarly spread between 25 and 45, with a slight concentration of participants in the late 20s, consistent with the mean of 34.2 years reported in the summary above.

### - Categorical Variables

The categorical variables include `treatment` (the primary predictor, with three levels: A, B, C, `sex`, and `PR` (the binary outcome).

```{r cat-dist}
p3 <- ggplot(df, aes(x = treatment, fill = treatment)) +
  geom_bar() +
  geom_text(stat = "count", aes(label = after_stat(count)), vjust = -0.5) +
  theme_bw() + theme(legend.position = "none") +
  labs(title = "Treatment group counts")

p4 <- ggplot(df, aes(x = sex, fill = sex)) +
  geom_bar() +
  geom_text(stat = "count", aes(label = after_stat(count)), vjust = -0.5) +
  theme_bw() + theme(legend.position = "none") +
  labs(title = "Sex distribution")

p5 <- ggplot(df, aes(x = PR, fill = PR)) +
  geom_bar() +
  geom_text(stat = "count", aes(label = after_stat(count)), vjust = -0.5) +
  theme_bw() + theme(legend.position = "none") +
  labs(title = "PR distribution")

p3 + p4 + p5
```

According to the distribution plots above, treatment group C has slightly more participants (n = 39) compared to A (n = 32) and B (n = 29). Sex is nearly equal between Female (n = 51) and Male (n = 49). The PR outcome is also evenly split (no: 49, yes: 51), indicating no class imbalance issue for binary modelling.

## · Descriptive statistics by treatment group

To compare the distribution of key variables across treatment groups, we generate a Table 1 stratified by treatment.

```{r table-one}
vars     <- c("age", "sex", "AF", "PR")
cat_vars <- c("sex", "PR")

tbl <- CreateTableOne(
  vars       = vars,
  strata     = "treatment",
  data       = df,
  factorVars = cat_vars
)

print(tbl, showAllLevels = TRUE, quote = FALSE, noSpaces = TRUE)
```

Treatment groups differ significantly in AF (p \< 0.001), with group B having the highest mean AF (76.93, SD = 11.22), followed by group C (69.24, SD = 13.00) and group A (37.53, SD = 6.37), suggesting a strong treatment effect. Age also differs marginally across groups (p = 0.040), with group B being slightly older on average (36.32 years). Sex distribution (p = 0.615) and PR rate (p = 0.084) do not differ significantly across treatment groups.

```{r kruskal-test}
kruskal.test(AF ~ treatment, data = df)
```

## · Correlation Matrix

To assess potential multicollinearity among continuous and genetic variables, we examine pairwise correlations between age, AF, and all SNP variables first using `corrplot`.

```{r corrplot, fig.height=15}
snp_cols <- names(df)[str_detect(names(df), "^rs")]

cor_data <- df |>
  select(age, AF, all_of(snp_cols)) |>
  mutate(across(everything(), as.numeric)) |>
  na.omit()

corrplot(cor(cor_data),
         method = "color",
         type   = "upper",
         addCoef.col = "black",
         tl.cex = 0.7,
         number.cex = 0.6,
         title  = "Correlation Matrix: Continuous & SNP Variables",
         mar    = c(0,0,1,0))
```

The correlation matrix reveals that most SNP variables show weak pairwise correlations with each other and with age and AF, suggesting limited multicollinearity among the genetic predictors. A small number of SNP pairs show moderate correlations (\|r\| \> 0.3), which may warrant attention during modelling.

# Primary Relationships / Plots

Having explored the structure and distribution of the data, we now examine the key relationships relevant to the research question: the effect of treatment on AF and PR, and the potential influence of covariates including age, sex, and genetic variables.

Smoking status and rs123456 are excluded from further analysis as both are zero-variance variables — all participants are non-smokers and all carry the CC genotype at rs123456, meaning neither variable can contribute any information to the model.

## Main topic: The impact of treatment on AF and PR

We first examine the primary relationship of interest — how treatment group relates to both the continuous outcome AF and the binary outcome PR.

```{r af-treatment, fig.cap="Distribution of AF by treatment group."}
ggplot(df, aes(x = treatment, y = AF, fill = treatment)) +
  geom_boxplot(alpha = 0.7, outlier.shape = 21) +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "AF by Treatment Group", x = "Treatment", y = "AF") +
  theme_bw() +
  theme(legend.position = "none")
```

```{r af-treatment-ppt}
ggplot(df, aes(x = treatment, y = AF, fill = treatment)) +
  geom_boxplot(alpha = 0.8, outlier.shape = 21, width = 0.5) +
  stat_summary(fun = mean, geom = "point", shape = 23,
               size = 4, fill = "white", colour = "black") +
  scale_fill_manual(values = trt_colours) +
  labs(title = "AF by Treatment Group",
       subtitle = "Diamond = group mean  ·  Kruskal–Wallis p < 0.001",
       x = "Treatment", y = "AF") +
  theme_ppt +
  theme(legend.position = "none")

save_ppt_fig("ppt_af_by_treatment.png")
```

AF values differ substantially across treatment groups. Group B shows the highest median AF (approximately 78), followed by group C (approximately 64), while group A has considerably lower values (median approximately 37) with a much tighter distribution. The non-overlapping interquartile ranges between group A and the other two groups strongly suggest a clinically meaningful treatment effect on AF.

```{r pr-treatment, fig.cap="Proportion of PR by treatment group."}
df |>
  filter(!is.na(PR)) |>
  group_by(treatment, PR) |>
  summarise(n = n(), .groups = "drop") |>
  mutate(prop = n / sum(n), .by = treatment) |>
  ggplot(aes(x = treatment, y = prop, fill = PR)) +
  geom_col(position = "dodge", alpha = 0.85) +
  scale_fill_brewer(palette = "Set1") +
  scale_y_continuous(labels = scales::percent) +
  labs(title = "PR Rate by Treatment Group", x = "Treatment",
       y = "Proportion", fill = "PR") +
  theme_bw()
```

```{r pr-treatment-ppt}
df |>
  filter(!is.na(PR)) |>
  group_by(treatment, PR) |>
  summarise(n = n(), .groups = "drop") |>
  mutate(prop = n / sum(n), .by = treatment) |>
  ggplot(aes(x = treatment, y = prop, fill = PR)) +
  geom_col(position = "dodge", alpha = 0.85, width = 0.6) +
  geom_text(aes(label = percent(prop, accuracy = 1)),
            position = position_dodge(0.6), vjust = -0.4, size = 5) +
  scale_fill_manual(values = c(no = "#B0C4CC", yes = "#0D7680"),
                    labels = c("PR: No", "PR: Yes")) +
  scale_y_continuous(labels = percent, limits = c(0, 0.85)) +
  labs(title = "PR Rate by Treatment Group",
       subtitle = "B: 65.5%  ·  A: 53.1%  ·  C: 38.5%  ·  χ² p = 0.084",
       x = "Treatment", y = "Proportion", fill = NULL) +
  theme_ppt

save_ppt_fig("ppt_pr_by_treatment.png")
```

Treatment group B has the highest PR rate (approximately 66%), followed by group A (approximately 53%), while group C has the lowest PR rate (approximately 38%) despite having the second highest AF values. This suggests that the relationship between treatment, AF, and PR is not straightforward, and that treatment C may produce higher AF values without necessarily translating to a PR.

## Sub-topic: The impact of Age on AF and PR

Age is a potential covariate that may confound the relationship between treatment and the outcomes. We examine whether age is associated with either AF or PR before formal modelling.

```{r af-age-marginal, fig.cap="Relationship between age and AF."}
ggplot(df, aes(x = age, y = AF)) +
  geom_point(alpha = 0.4, colour = "steelblue") +
  geom_smooth(method = "lm", se = TRUE, colour = "red") +
  labs(title = "AF vs Age", x = "Age", y = "AF") +
  theme_bw()
```

```{r af-age-marginal-ppt}
ggplot(df, aes(x = age, y = AF)) +
  geom_point(alpha = 0.45, colour = "#0D7680", size = 2.5) +
  geom_smooth(method = "lm", se = TRUE, colour = "#E8A838", fill = "#E8A838", alpha = 0.15) +
  labs(title = "AF vs Age",
       subtitle = "Positive trend; considerable scatter driven by treatment differences",
       x = "Age (years)", y = "AF") +
  theme_ppt

save_ppt_fig("ppt_af_vs_age.png", width = 8, height = 6)
```

There is a positive linear relationship between age and AF, with older participants tending to have higher AF values. However, there is considerable scatter around the regression line, suggesting that age alone does not fully explain the variation in AF. This indicates that age may act as a covariate worth adjusting for in formal modelling, but treatment group is likely the dominant predictor.

```{r pr-age, fig.cap="Age distribution by PR outcome."}
ggplot(df |> filter(!is.na(PR)), aes(x = PR, y = age, fill = PR)) +
  geom_boxplot(alpha = 0.7) +
  scale_fill_brewer(palette = "Set1") +
  labs(title = "Age by PR Outcome", x = "PR", y = "Age") +
  theme_bw() +
  theme(legend.position = "none")
```

```{r pr-age-ppt}
ggplot(df |> filter(!is.na(PR)), aes(x = PR, y = age, fill = PR)) +
  geom_boxplot(alpha = 0.8, width = 0.45) +
  scale_fill_manual(values = c(no = "#B0C4CC", yes = "#0D7680")) +
  labs(title = "Age by PR Outcome",
       subtitle = "Median: 29 yrs (PR = no)  vs  40 yrs (PR = yes)",
       x = "PR", y = "Age (years)") +
  theme_ppt +
  theme(legend.position = "none")

save_ppt_fig("ppt_age_by_pr.png", width = 8, height = 6)
```

Participants with a PR (PR = yes) tend to be considerably older (median approximately 40 years) compared to those without (median approximately 29 years), with minimal overlap between the interquartile ranges of the two groups. This suggests that age is strongly associated with PR outcome and should be included as a covariate in any model predicting PR.

## Sub-topic: The impact of sex on AF and PR

We next examine whether sex is associated with AF or PR, as it may act as a confounding variable in the relationship between treatment and the outcomes.

```{r af-sex, fig.cap="AF by sex."}
ggplot(df, aes(x = sex, y = AF, fill = sex)) +
  geom_boxplot(alpha = 0.7) +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "AF by Sex", x = "Sex", y = "AF") +
  theme_bw() +
  theme(legend.position = "none")
```

```{r af-sex-ppt}
ggplot(df, aes(x = sex, y = AF, fill = sex)) +
  geom_boxplot(alpha = 0.8, width = 0.45) +
  stat_summary(
    fun = median,
    geom = "text",
    aes(label = round(after_stat(y), 1)),
    vjust = -0.5,
    size = 5,
    fontface = "bold"
  ) +
  scale_fill_manual(values = c(Female = "#7FB3C8", Male = "#4A7FA5")) +
  labs(title = "AF by Sex",
       subtitle = "Distributions are similar — sex unlikely to confound AF",
       x = "Sex", y = "AF") +
  theme_ppt +
  theme(legend.position = "none")
save_ppt_fig("ppt_af_by_sex.png", width = 8, height = 6)
```

AF distributions are similar between females and males, with both groups having comparable medians (approximately 60 and 63 respectively) and largely overlapping interquartile ranges. This suggests that sex has little marginal association with AF, though it may still be worth adjusting for in formal modelling.

```{r pr-sex, fig.cap="PR rate by sex."}
df |>
  filter(!is.na(PR), !is.na(sex)) |>
  group_by(sex, PR) |>
  summarise(n = n(), .groups = "drop") |>
  mutate(prop = n / sum(n), .by = sex) |>
  ggplot(aes(x = sex, y = prop, fill = PR)) +
  geom_col(position = "dodge", alpha = 0.85) +
  scale_fill_brewer(palette = "Set1") +
  scale_y_continuous(labels = scales::percent) +
  labs(title = "PR Rate by Sex", x = "Sex", y = "Proportion", fill = "PR") +
  theme_bw()
```

```{r pr-sex-ppt}
df |>
  filter(!is.na(PR), !is.na(sex)) |>
  group_by(sex, PR) |>
  summarise(n = n(), .groups = "drop") |>
  mutate(prop = n / sum(n), .by = sex) |>
  ggplot(aes(x = sex, y = prop, fill = PR)) +
  geom_col(position = "dodge", alpha = 0.85, width = 0.6) +
  geom_text(aes(label = percent(prop, accuracy = 1)),
            position = position_dodge(0.6), vjust = -0.4, size = 5) +
  scale_fill_manual(values = c(no = "#B0C4CC", yes = "#0D7680"),
                    labels = c("PR: No", "PR: Yes")) +
  scale_y_continuous(labels = percent, limits = c(0, 0.75)) +
  labs(title = "PR Rate by Sex",
       subtitle = "~51% in both sexes — no meaningful association with PR",
       x = "Sex", y = "Proportion", fill = NULL) +
  theme_ppt

save_ppt_fig("ppt_pr_by_sex.png", width = 8, height = 6)
```

The PR rate is virtually identical between females and males (both approximately 51% positive), indicating that sex has no meaningful association with PR in this dataset. Sex is therefore unlikely to be an important covariate for the PR outcome in formal modelling.

## Sub-topic: The effect of SNP on AF and PR

Before examining SNP–outcome associations, we first look at how genotypes are distributed across dosage groups. Most SNPs in this sample are dominated by heterozygous carriers (dosage = 1), with relatively few participants in the homozygous alternate group (dosage = 2) — meaning effect estimates at dosage 2 should be interpreted with caution due to small cell sizes.

```{r snp-dosage-example}
# Select two representative SNPs to illustrate the dosage distribution
example_snps <- c("rs19486", "rs91154")

df |>
  select(all_of(example_snps)) |>
  pivot_longer(everything(), names_to = "SNP", values_to = "dosage") |>
  filter(!is.na(dosage)) |>
  group_by(SNP, dosage) |>
  summarise(n = n(), .groups = "drop") |>
  mutate(pct = n / sum(n), .by = SNP) |>
  ggplot(aes(x = dosage, y = n, fill = dosage)) +
  geom_col(width = 0.55, alpha = 0.85) +
  geom_text(aes(label = paste0("n=", n, "\n(", round(pct*100), "%)")),
            vjust = -0.3, size = 4.5, fontface = "bold", lineheight = 0.9) +
  scale_fill_manual(values = c("0" = "#A8D5D9", "1" = "#0D7680", "2" = "#1B2B4B")) +
  scale_y_continuous(limits = c(0, 70), expand = c(0, 0)) +
  facet_wrap(~ SNP) +
  labs(
    title    = "Dosage Distribution — Example SNPs",
    subtitle = "Number of participants per alternate allele dosage group",
    x        = "Alternate Allele Dosage (0 / 1 / 2)",
    y        = "Count"
  ) +
  theme_ppt +
  theme(legend.position = "none")

save_ppt_fig("ppt_snp_dosage_example.png", width = 8, height = 7)
```

Finally, we examine whether genetic variation, as captured by the SNP variables, is associated with AF or PR. Each SNP is encoded under an additive model with dosage values of 0, 1, or 2 representing the number of alternate alleles carried by each participant.

```{r snp-af, fig.cap="AF distribution by SNP dosage for each SNP.", fig.height=10}
snp_cols <- names(df)[str_detect(names(df), "^rs") & 
                        names(df) != "rs123456"]  # 排除零方差列

df |>
  select(AF, all_of(snp_cols)) |>
  pivot_longer(-AF, names_to = "SNP", values_to = "dosage") |>
  filter(!is.na(dosage)) |>
  ggplot(aes(x = dosage, y = AF, fill = dosage)) +
  geom_boxplot(alpha = 0.7) +
  facet_wrap(~ SNP, scales = "free_x") +
  scale_fill_brewer(palette = "Blues") +
  labs(title = "AF by SNP Dosage", x = "Alternate Allele Dosage",
       y = "AF") +
  theme_bw() +
  theme(legend.position = "none")
```

Across most SNPs, the distribution of AF appears similar across dosage groups (0, 1, 2), with largely overlapping interquartile ranges, suggesting that the majority of SNPs have little marginal association with AF. However, a small number of SNPs show modest differences across dosage groups — for example, rs88128 appear to show a trend of increasing and rs19486 showing a trend of decreasing AF with higher alternate allele dosage. These SNPs may be worth investigating further in formal modelling, though the small sample sizes within each dosage group warrant caution in interpretation.

To identify which SNPs show the strongest marginal association with AF, we computed the Spearman correlation between each SNP's alternate allele dosage and AF across all 50 SNPs. The top 10 SNPs ranked by \|r\| are shown below, with the direction of association indicated by colour.

```{r snp-top10-af}
snp_cols_clean <- names(df)[str_detect(names(df), "^rs") & names(df) != "rs123456"]

# Calculate the Spearman’s correlation coefficient between each SNP and AF
snp_cor_af <- tibble(
  SNP = snp_cols_clean,
  r   = sapply(snp_cols_clean, function(s) {
    cor(as.numeric(df[[s]]), df$AF, use = "complete.obs", method = "spearman")
  })
) |>
  arrange(desc(abs(r))) |>
  slice_head(n = 10) |>
  mutate(
    SNP       = fct_reorder(SNP, abs(r)),
    direction = if_else(r > 0, "Positive", "Negative")
  )

ggplot(snp_cor_af, aes(x = abs(r), y = SNP, fill = direction)) +
  geom_col(width = 0.65, alpha = 0.85) +
  geom_text(aes(label = sprintf("r = %.3f", r)),
            hjust = -0.1, size = 4.5, fontface = "bold") +
  scale_fill_manual(values = c(Positive = "#0D7680", Negative = "#E8A838")) +
  scale_x_continuous(limits = c(0, 0.55), expand = c(0, 0)) +
  labs(
    title    = "Top 10 SNPs Associated with AF",
    subtitle = "Spearman correlation  ·  sorted by |r|  ·  exploratory only",
    x        = "|r|  (Spearman)",
    y        = NULL,
    fill     = "Direction"
  ) +
  theme_ppt +
  theme(legend.position = "right")

save_ppt_fig("ppt_snp_top10_af.png", width = 8, height = 6)
```

All top 10 correlations are weak (\|r\| ≤ 0.23), consistent with the expectation that individual SNPs rarely explain large proportions of variance in complex traits. rs19486 and rs26922 show the strongest negative associations with AF, while rs88128 shows the strongest positive association. These effect sizes are modest and should be interpreted cautiously — formal testing with appropriate multiple testing correction (e.g. Bonferroni or FDR) will be required before drawing any conclusions about individual SNP effects.

To complement the AF analysis, we examine whether SNP dosage is associated with PR outcome. For each SNP, the proportion of participants with PR = yes and PR = no is plotted separately for each dosage group (0, 1, 2).

```{r snp-pr, fig.cap="PR rate by SNP dosage for each SNP.", fig.height=10}
df |>
  select(PR, all_of(snp_cols)) |>
  pivot_longer(-PR, names_to = "SNP", values_to = "dosage") |>
  filter(!is.na(dosage), !is.na(PR)) |>
  group_by(SNP, dosage, PR) |>
  summarise(n = n(), .groups = "drop") |>
  mutate(prop = n / sum(n), .by = c(SNP, dosage)) |>
  ggplot(aes(x = dosage, y = prop, fill = PR)) +
  geom_col(position = "dodge", alpha = 0.85) +
  facet_wrap(~ SNP) +
  scale_fill_brewer(palette = "Set1") +
  scale_y_continuous(labels = scales::percent) +
  labs(title = "PR Rate by SNP Dosage", x = "Alternate Allele Dosage",
       y = "Proportion", fill = "PR") +
  theme_bw()
```

Similar to the AF results, most SNPs show little variation in PR rate across dosage groups, with the proportion of positive responses remaining roughly 50/50 regardless of alternate allele dosage. A few SNPs such as rs91154 and rs58943 show more noticeable shifts in PR rate across dosage levels, which may warrant closer examination in formal modelling. Overall, the marginal effect of individual SNPs on PR appears modest at this exploratory stage.

Similarly, to identify which SNPs show the strongest marginal association with PR, we computed the Spearman correlation between each SNP's alternate allele dosage and PR outcome (coded as 0 = no, 1 = yes) across all 50 SNPs. The top 10 SNPs ranked by \|r\| are shown below.

```{r snp-top10-pr}
# Converting PR to 0/1 values
pr_numeric <- as.numeric(df$PR) - 1  # no=0, yes=1

snp_cor_pr <- tibble(
  SNP = snp_cols_clean,
  r   = sapply(snp_cols_clean, function(s) {
    cor(as.numeric(df[[s]]), pr_numeric, use = "complete.obs", method = "spearman")
  })
) |>
  arrange(desc(abs(r))) |>
  slice_head(n = 10) |>
  mutate(
    SNP       = fct_reorder(SNP, abs(r)),
    direction = if_else(r > 0, "Positive", "Negative")
  )

ggplot(snp_cor_pr, aes(x = abs(r), y = SNP, fill = direction)) +
  geom_col(width = 0.65, alpha = 0.85) +
  geom_text(aes(label = sprintf("r = %.3f", r)),
            hjust = -0.1, size = 4.5, fontface = "bold") +
  scale_fill_manual(values = c(Positive = "#0D7680", Negative = "#E8A838")) +
  scale_x_continuous(limits = c(0, 0.55), expand = c(0, 0)) +
  labs(
    title    = "Top 10 SNPs Associated with PR",
    subtitle = "Spearman correlation (SNP dosage vs binary PR)  ·  exploratory only",
    x        = "|r|  (Spearman)",
    y        = NULL,
    fill     = "Direction"
  ) +
  theme_ppt +
  theme(legend.position = "right")

#save_ppt_fig("ppt_snp_top10_pr.png", width = 8, height = 6)
```

rs91154 shows the strongest negative association with PR (r = −0.267), meaning higher alternate allele dosage is associated with a lower probability of PR. Nine of the top 10 SNPs show negative associations, with rs78356 being the only exception. As with AF, all correlations remain weak (\|r\| ≤ 0.27), and these findings are exploratory — formal testing with multiple testing correction is required before any SNP can be considered a candidate predictor of PR.

To visualise the dosage-response relationship more directly, we plot the observed PR probability at each dosage level (0, 1, 2) for the top 10 SNPs identified above. A downward trend across dosage groups would be consistent with the negative correlations observed, suggesting that carrying more alternate alleles is associated with a lower probability of PR.

```{r snp-dosage-pr-ppt}
# Using the top 10 PR SNPs calculated earlier
top_pr_snps <- snp_cor_pr |> pull(SNP) |> as.character()

df |>
  select(PR, all_of(top_pr_snps)) |>
  pivot_longer(-PR, names_to = "SNP", values_to = "dosage") |>
  filter(!is.na(dosage), !is.na(PR)) |>
  group_by(SNP, dosage) |>
  summarise(
    prob = mean(PR == "yes"),
    n    = n(),
    .groups = "drop"
  ) |>
  mutate(dosage = as.numeric(as.character(dosage))) |>
  ggplot(aes(x = dosage, y = prob, colour = SNP, group = SNP)) +
  geom_line(linewidth = 0.9, alpha = 0.8) +
  geom_point(aes(size = n), alpha = 0.85) +
  scale_y_continuous(labels = percent, limits = c(0, 1)) +
  scale_x_continuous(breaks = c(0, 1, 2),
                     labels = c("0\n(hom ref)", "1\n(het)", "2\n(hom alt)")) +
  scale_size_continuous(range = c(2, 6), guide = "none") +
  labs(
    title    = "PR Probability by SNP Dosage — Top 10 SNPs",
    subtitle = "Point size proportional to n",
    x        = "Alternate Allele Dosage",
    y        = "PR Probability",
    colour   = "SNP"
  ) +
  theme_ppt +
  theme(legend.position = "right")

#save_ppt_fig("ppt_snp_dosage_pr_prob.png", width = 9.3, height = 7)
```

The majority of SNPs show a consistent downward trend from dosage 0 to dosage 2, corroborating the negative correlations observed above. rs91154 shows the steepest decline, dropping from approximately 75% PR probability at dosage 0 to around 38% at dosage 2. rs78356 is the clear exception, showing an upward trend consistent with its positive correlation. Note that dosage 2 points are based on small cell sizes for most SNPs — the apparent trends at the homozygous alternate level should be interpreted with caution until confirmed in formal modelling.

To provide a broader overview across all 50 SNPs, we plot PR probability by dosage for every SNP simultaneously, highlighting the 10 SNPs with the highest observed PR probability at any dosage level. Grey lines represent the remaining SNPs as background context.

```{r snp-dosage-pr-prob-all-ppt}
snp_probs <- df |>
  select(PR, all_of(snp_cols_clean)) |>
  pivot_longer(-PR, names_to = "SNP", values_to = "dosage") |>
  filter(!is.na(dosage), !is.na(PR)) |>
  group_by(SNP, dosage) |>
  summarise(
    prob = mean(PR == "yes"),
    n    = n(),
    .groups = "drop"
  ) |>
  mutate(dosage = as.numeric(as.character(dosage)))

# Identify the top 10 SNPs with the highest probability of PR at any given dosage
top10_snps <- snp_probs |>
  group_by(SNP) |>
  summarise(max_prob = max(prob)) |>
  arrange(desc(max_prob)) |>
  slice_head(n = 10) |>
  pull(SNP)

snp_probs |>
  mutate(
    highlight = if_else(SNP %in% top10_snps, SNP, "other"),
    alpha_val = if_else(SNP %in% top10_snps, 1, 0.15),
    lwd_val   = if_else(SNP %in% top10_snps, 1.1, 0.5)
  ) |>
  ggplot(aes(x = dosage, y = prob, group = SNP,
             colour = highlight, alpha = alpha_val, linewidth = lwd_val)) +
  geom_line() +
  geom_point(data = ~ filter(., SNP %in% top10_snps),
             aes(size = n), alpha = 1) +
  scale_colour_manual(
    values = c(setNames(scales::hue_pal()(10), top10_snps), other = "grey70"),
    breaks = top10_snps   # legend: Show top 10 only
  ) +
  scale_alpha_identity() +
  scale_linewidth_identity() +
  scale_size_continuous(range = c(2, 5), guide = "none") +
  scale_y_continuous(labels = percent, limits = c(0, 1)) +
  scale_x_continuous(breaks = c(0, 1, 2),
                     labels = c("0\n(hom ref)", "1\n(het)", "2\n(hom alt)")) +
  labs(
    title    = "PR Probability by SNP Dosage — All SNPs",
    subtitle = "Top 10 by max PR probability highlighted",
    x        = "Alternate Allele Dosage",
    y        = "PR Probability",
    colour   = "Top 10 SNPs"
  ) +
  theme_ppt +
  theme(legend.position = "right")

save_ppt_fig("ppt_snp_dosage_pr_prob_all.png", width = 9.3, height = 7)
```

Unlike the top 10 by correlation (which were predominantly negative), the top 10 by maximum PR probability show more heterogeneous patterns — some lines trend downward (e.g. rs91154), while others increase sharply at dosage 2 (e.g. rs56579, rs78356). This divergence highlights that high maximum probability does not necessarily imply a consistent dosage-response relationship. The crossing of lines at dosage 1 and the wide spread at dosage 2 again reflect the small sample sizes in the homozygous alternate group, which can produce unstable probability estimates. These SNPs warrant closer inspection in formal modelling, but no conclusions should be drawn from exploratory plots alone.

## Interaction：Treatment × Age → AF

Given that both treatment and age appear to be associated with AF, we examine whether the effect of age on AF differs across treatment groups, which would indicate a treatment-age interaction.

```{r af-age-treatment, fig.cap="AF vs age stratified by treatment group."}
ggplot(df, aes(x = age, y = AF, colour = treatment)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = TRUE) +
  scale_colour_brewer(palette = "Set2") +
  labs(title = "AF vs Age by Treatment Group", x = "Age", y = "AF",
       colour = "Treatment") +
  theme_bw()
```

```{r af-age-treatment-ppt}
ggplot(df, aes(x = age, y = AF, colour = treatment)) +
  geom_point(alpha = 0.5, size = 2.5) +
  geom_smooth(method = "lm", se = TRUE, alpha = 0.12) +
  scale_colour_manual(values = trt_colours) +
  labs(title = "AF vs Age by Treatment Group",
       subtitle = "Parallel slopes → no treatment × age interaction; treatment is the dominant predictor",
       x = "Age (years)", y = "AF", colour = "Treatment") +
  theme_ppt

save_ppt_fig("ppt_af_age_by_treatment.png", width = 9, height = 6)
```

When stratified by treatment, the relationship between age and AF becomes remarkably strong within each group, with points clustering tightly around the regression lines. The three lines run parallel with similar slopes, confirming that the effect of age on AF is consistent across treatment groups with no evidence of a treatment-age interaction. The large vertical separation between group A and groups B and C confirms that the wide scatter seen in the unstratified plot was largely driven by treatment differences, reinforcing that treatment is a dominant predictor of AF that must be accounted for in any formal model.

## Interaction：Treatment × Sex → AF

We also examine whether the effect of treatment on AF differs by sex, which would indicate a treatment-sex interaction.

```{r af-sex-treatment, fig.cap="AF by treatment group, stratified by sex."}
ggplot(df, aes(x = treatment, y = AF, fill = sex)) +
  geom_boxplot(alpha = 0.7, position = position_dodge(0.8)) +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "AF by Treatment and Sex", x = "Treatment", y = "AF",
       fill = "Sex") +
  theme_bw()
```

```{r af-sex-treatment-ppt}
ggplot(df, aes(x = treatment, y = AF, fill = sex)) +
  geom_boxplot(alpha = 0.8, position = position_dodge(0.8)) +
  scale_fill_manual(values = c(Female = "#7FB3C8", Male = "#4A7FA5")) +
  labs(title = "AF by Treatment and Sex",
       subtitle = "Overlapping IQRs within each group — little evidence of treatment × sex interaction",
       x = "Treatment", y = "AF", fill = "Sex") +
  theme_ppt

save_ppt_fig("ppt_af_treatment_by_sex.png", width = 9, height = 6)
```

Within each treatment group, AF distributions are largely similar between females and males, with overlapping interquartile ranges across all three groups. The dominant pattern remains the treatment effect — group A has substantially lower AF than groups B and C regardless of sex. There is a slight tendency for males to have marginally higher AF in groups A and C, but the overlap is considerable and this difference appears negligible. Overall, there is little evidence of a treatment-sex interaction, suggesting that the effect of treatment on AF does not differ meaningfully between females and males.

# Models

Based on EDA findings, 2 separate models are fitted to address the 2 outcomes of interest:

1.  **AF (continuous): multiple linear regression**

AF is a continuous outcome with no natural upper or lower bound in the context of this study. Linear regression is the appropriate choice because it models the expected value of a continuous response as a linear combination of predictors, and the EDA showed that AF follows a roughly symmetric distribution within each treatment group with no severe skew. The model estimates how much AF changes, on average, for a one-unit change in each predictor — making the coefficients directly interpretable as differences in AF values.

2.  **PR (binary): logistic regression**

PR is a binary outcome: each participant either had a positive response (yes) or did not (no). Standard linear regression is unsuitable here because it can produce predicted probabilities outside the range [0, 1] and assumes a linear relationship between predictors and the outcome on the raw scale. Logistic regression instead models the log-odds of PR = yes as a linear function of the predictors, which guarantees that all predicted probabilities fall between 0 and 1. Coefficients are exponentiated to produce odds ratios, which express how the odds of a positive response change with each predictor.

Further details on model selection and interpretation are provided in the subsections below.

Both models follow the same structure: a baseline model with treatment only, followed by a main-effects model adding age and sex as covariates, and finally a SNP-augmented model incorporating the top genetic candidates identified during EDA. Smoking status and `rs123456` are excluded from all models as zero-variance variables.

## · Choose SNP candidates

As a general rule of thumb, reliable regression estimates require limited observations per predictor, including all 51 SNPs at once would far exceed this limit, leading to unstable and unreliable coefficient estimates. Therefore, univariable regression is used at the screening stage — rather than correlation alone — because it places each SNP on the same inferential scale as the final model, producing a p-value that reflects the significance of the SNP–outcome association under the same modelling framework.

For each of the 50 SNPs, a separate regression is fitted with just that SNP as the sole predictor, one at a time.

-   For AF, linear regression is used;
-   for PR, logistic regression.

This gives a p-value for each SNP indicating whether it has a statistically significant association with the outcome on its own.

Two thresholds are checked:

-   the stricter **Bonferroni threshold** (0.05 / 50 = 0.001), which accounts for the fact that we are running 50 tests at once and would expect some false positives by chance, and
-   the more lenient **nominal threshold** (p \< 0.05).

SNPs passing the nominal threshold are taken forward into the multivariable models. If few or no SNPs survive the stricter Bonferroni correction, which is expected given the small sample size, the nominal threshold is used instead, and any SNP findings are treated as preliminary rather than definitive.

```{r snp-screen}
snp_cols_model <- names(df)[str_detect(names(df), "^rs") & names(df) != "rs123456"]

# AF: univariable linear regression for each SNP
snp_screen_af <- tibble(
  SNP  = snp_cols_model,
  p    = sapply(snp_cols_model, function(s) {
    summary(lm(AF ~ as.numeric(df[[s]]), data = df))$coefficients[2, 4]
  }),
  beta = sapply(snp_cols_model, function(s) {
    coef(lm(AF ~ as.numeric(df[[s]]), data = df))[2]
  })
) |>
  arrange(p) |>
  mutate(bonferroni = p < 0.05 / n(),          # Bonferroni-corrected threshold
         nominal    = p < 0.05)

cat("SNPs passing Bonferroni threshold (AF):", sum(snp_screen_af$bonferroni), "\n")
cat("SNPs passing nominal p < 0.05 (AF):",    sum(snp_screen_af$nominal),    "\n")
print(head(snp_screen_af, 10))

# PR: univariable logistic regression for each SNP
pr_num <- as.numeric(df$PR) - 1   # recode to 0/1

snp_screen_pr <- tibble(
  SNP  = snp_cols_model,
  p    = sapply(snp_cols_model, function(s) {
    summary(glm(pr_num ~ as.numeric(df[[s]]),
                family = binomial))$coefficients[2, 4]
  }),
  OR   = sapply(snp_cols_model, function(s) {
    exp(coef(glm(pr_num ~ as.numeric(df[[s]]),
                 family = binomial))[2])
  })
) |>
  arrange(p) |>
  mutate(bonferroni = p < 0.05 / n(),
         nominal    = p < 0.05)

cat("\nSNPs passing Bonferroni threshold (PR):", sum(snp_screen_pr$bonferroni), "\n")
cat("SNPs passing nominal p < 0.05 (PR):",    sum(snp_screen_pr$nominal),    "\n")
print(head(snp_screen_pr, 10))
```

No SNPs passed the Bonferroni-corrected threshold for either outcome, which is unsurprising given the small sample size of 100 participants.

```{r snp-candidates}
# Retain SNPs that pass the nominal threshold (p < 0.05) for inclusion in the augmented models. Since none pass Bonferroni, we use nominal significance and note this as exploratory.
top_snps_af <- snp_screen_af |> filter(nominal) |> pull(SNP)
top_snps_pr <- snp_screen_pr |> filter(nominal) |> pull(SNP)

cat("Top AF SNP candidates:", paste(top_snps_af, collapse = ", "), "\n")
cat("Top PR SNP candidates:", paste(top_snps_pr, collapse = ", "), "\n")
```

For AF, one SNP passed the nominal threshold: **rs19486** (p = 0.015, β = −6.23), indicating that each additional alternate allele is associated with a decrease of 6.23 units in AF on average.

For PR, one SNP passed the nominal threshold: **rs91154** (p = 0.010, OR = 0.438), suggesting that each additional alternate allele is associated with lower odds of a positive response.

Both findings are consistent with the EDA results observed earlier. These two SNPs are carried forward as candidates into the multivariable models, and their effects should be interpreted as exploratory given that they did not survive Bonferroni correction.

## · AF Model

Three linear regression models are fitted for AF, each building on the previous one.

1.  The **baseline model** includes treatment only, to capture the raw treatment effect seen in the EDA.

2.  The **main effects model** adds age and sex as covariates, allowing us to estimate the treatment effect after accounting for any differences in participant characteristics across groups.

3.  The **third model** further adds rs19486, the one SNP that passed the nominal screening threshold, to assess whether genetic variation explains any additional variation in AF beyond treatment, age, and sex.

Comparing these three models allows us to see how much each additional set of variables improves the fit, and whether the treatment effect changes once other factors are accounted for.

```{r af-model-fit}

df_model <- df |> drop_na(age)   # Remove NA rows in Age

# Model 0: intercept only (for reference R²)
lm_af_null <- lm(AF ~ 1, data = df_model)
cat("Model 0: Intercept only\n")
print(lm_af_null)

# Model 1: treatment only
lm_af_base <- lm(AF ~ treatment, data = df_model)
cat("-------\n")
cat("Model 1: Treatment only\n")
print(lm_af_base)

# Model 2: treatment + age + sex
lm_af_main <- lm(AF ~ treatment + age + sex, data = df_model)
cat("-------\n")
cat("Model 2: Treatment + age + sex\n")
print(lm_af_main)

# Model 3: treatment + age + sex + top SNP candidates
# Build the formula dynamically from the screened candidates
if (length(top_snps_af) > 0) {
  snp_terms_af <- paste(
    sprintf("as.numeric(%s)", top_snps_af),
    collapse = " + "
  )
  lm_af_snp <- lm(
    as.formula(paste("AF ~ treatment + age + sex +", snp_terms_af)),
    data = df
  )
  cat("-------\n")
  cat("Model 3: Treatment + age + sex +", paste(top_snps_af, collapse = ", "), "\n")
} else {
  lm_af_snp <- lm_af_main
  cat("-------\n")
  cat("Model 3: No SNPs passed nominal screening — falling back to Model 2\n")
}
print(lm_af_snp)
```

**Model 0 (Intercept only)** estimates the overall mean AF across all participants as 61.32, with no predictors included. This serves as a baseline reference for comparing model fit.

**Model 1 (Treatment only)** shows that treatment group A has a mean AF of 37.53. Groups B and C have substantially higher AF values, with B averaging 39.40 units higher and C averaging 31.71 units higher than A. This confirms the strong treatment effect observed in the EDA.

**Model 2 (Treatment + age + sex)** retains the large treatment effects after adjusting for age and sex, with B and C remaining 35.72 and 34.34 units above A respectively. Age has a positive association with AF (β = 1.65), meaning each additional year of age is associated with an increase of 1.65 units in AF on average. Sex has a negligible effect (β = 0.71).

**Model 3 (Treatment + age + sex + rs19486)** produces very similar coefficient estimates to Model 2, with rs19486 contributing a near-zero additional effect (β = 0.26). This suggests that rs19486 adds little explanatory value for AF beyond the other predictors already in the model, despite passing the nominal screening threshold.

## · PR Model

Three logistic regression models are fitted for PR, following the same structure as the AF models. 

1. The **baseline model** includes treatment only, to capture the raw association between treatment group and the probability of a positive response. 

2. The **main effects model** adds age and sex, given that the EDA showed a strong association between age and PR outcome. 

3. The **third model** further adds rs91154, the one SNP that passed the nominal screening threshold for PR, to assess whether it explains any additional variation in the probability of a positive response beyond treatment, age, and sex.

```{r pr-model-fit}
# Model 1: treatment only
glm_pr_base <- glm(PR ~ treatment, data = df_model, family = binomial)
cat("Model 1: Treatment only\n")
print(glm_pr_base)

# Model 2: treatment + age + sex
glm_pr_main <- glm(PR ~ treatment + age + sex, data = df_model, family = binomial)
cat("-------\n")
cat("Model 2: Treatment + age + sex\n")
print(glm_pr_main)

# Model 3: treatment + age + sex + top SNP candidates
if (length(top_snps_pr) > 0) {
  snp_terms_pr <- paste(
    sprintf("as.numeric(%s)", top_snps_pr),
    collapse = " + "
  )
  glm_pr_snp <- glm(
    as.formula(paste("PR ~ treatment + age + sex +", snp_terms_pr)),
    data   = df,
    family = binomial
  )
  cat("-------\n")
  cat("Model 3: Treatment + age + sex +", paste(top_snps_pr, collapse = ", "), "\n")
} else {
  glm_pr_snp <- glm_pr_main
  cat("-------\n")
  cat("Model 3: No SNPs passed nominal screening — falling back to Model 2\n")
}
print(glm_pr_snp)
```

**Model 1 (Treatment only)** shows that treatment B is associated with higher log-odds of PR compared to A (β = 0.52), while treatment C is associated with lower log-odds (β = −0.60). However, the small reduction in deviance from 138.6 to 133.6 suggests that treatment alone explains relatively little variation in PR.

**Model 2 (Treatment + age + sex)** shows a substantial improvement in fit, with residual deviance dropping from 133.6 to 93.9 and AIC improving from 139.6 to 103.9. This confirms that age is a strong predictor of PR (β = 0.26), consistent with the EDA finding that participants with PR = yes tended to be considerably older. Notably, after adjusting for age, the treatment effect becomes much smaller and no longer appears meaningful, suggesting that the raw treatment differences in PR seen in the EDA were partly driven by age differences across groups.

**Model 3 (Treatment + age + sex + rs91154)** shows a further modest improvement in fit (AIC: 102.6 vs 103.9). rs91154 has a negative coefficient (β = −0.71), indicating that each additional alternate allele is associated with lower log-odds of PR, consistent with the EDA screening result. The treatment and age coefficients remain largely unchanged from Model 2, suggesting rs91154 contributes some independent explanatory value for PR.

# Explore models / interpret models

## · AF Model

```{r af-model-compare}
# Compare models using AIC and adjusted R²
model_compare_af <- tibble(
  Model       = c("Null", "Treatment only", "Main effects", "Main + SNPs"),
  AIC         = c(AIC(lm_af_null), AIC(lm_af_base),
                  AIC(lm_af_main), AIC(lm_af_snp)),
  Adj_R2      = c(
    summary(lm_af_null)$adj.r.squared,
    summary(lm_af_base)$adj.r.squared,
    summary(lm_af_main)$adj.r.squared,
    summary(lm_af_snp)$adj.r.squared
  )
) |>
  mutate(across(where(is.numeric), ~ round(.x, 3)))

print(model_compare_af)

# ANOVA to formally test whether adding covariates improves fit
anova(lm_af_base, lm_af_main, lm_af_snp)
```

Adding age and sex dramatically improves model fit (adjusted R² from 0.705 to 0.977, AIC drop of ~266 units; ANOVA p < 0.001). Adding rs19486 provides no further gain, so the main-effects model is preferred.

```{r af-model-summary}
# Full summary of the selected main-effects model
summary(lm_af_main)

# 95% confidence intervals for all coefficients
confint(lm_af_main)
```

After adjusting for age and sex, treatment B and C are associated with AF values approximately 35.7 and 34.3 units higher than A respectively (both p < 0.001). Each additional year of age increases AF by ~1.65 units (p < 0.001). Sex has no meaningful effect as we expected. The model explains 97.7% of variance in AF.

## · PR Model

```{r pr-model-compare}
# Compare models using AIC and McFadden's pseudo-R²
mcfadden_r2 <- function(model) {
  1 - (logLik(model) / logLik(update(model, . ~ 1)))
}

model_compare_pr <- tibble(
  Model        = c("Treatment only", "Main effects", "Main + SNPs"),
  AIC          = c(AIC(glm_pr_base), AIC(glm_pr_main), AIC(glm_pr_snp)),
  McFadden_R2  = c(
    as.numeric(mcfadden_r2(glm_pr_base)),
    as.numeric(mcfadden_r2(glm_pr_main)),
    as.numeric(mcfadden_r2(glm_pr_snp))
  )
) |>
  mutate(across(where(is.numeric), ~ round(.x, 3)))

print(model_compare_pr)

# Likelihood ratio test
anova(glm_pr_base, glm_pr_main, glm_pr_snp, test = "LRT")
```

The main-effects model substantially outperforms treatment alone (McFadden's R² from 0.036 to 0.322, AIC from 139.6 to 103.9; LRT p < 0.001). Adding rs91154 yields only a marginal further improvement (AIC: 102.6). The main-effects model (treatment + age + sex) is selected as the final PR model, as the addition of rs91154 yields only a marginal AIC improvement (103.9 → 102.6) and the SNP finding did not survive Bonferroni correction.

```{r pr-model-summary}
# Full summary of the selected main-effects model
summary(glm_pr_main)

# Odds ratios and 95% CIs (exponentiate log-odds coefficients)
or_table <- tidy(glm_pr_main, conf.int = TRUE, exponentiate = TRUE) |>
  filter(term != "(Intercept)") |>
  mutate(across(c(estimate, conf.low, conf.high, p.value), ~ round(.x, 4)))

print(or_table)
```

After adjusting for age and sex, neither treatment B nor C differs significantly from A in the odds of PR (both p > 0.05) — the raw treatment differences seen in the EDA were largely driven by age confounding. Age is the dominant predictor (OR ≈ 1.30 per year, p < 0.001). Sex shows no meaningful association as we expected.

# Assumption Checking

## · AF: Linear Regression Diagnostics

Four standard diagnostic plots are examined to assess whether the linear regression assumptions are satisfied: linearity, homoscedasticity (constant variance), normality of residuals, and absence of highly influential observations.

```{r af-diagnostics, fig.height=8, fig.width=10}
# Base R diagnostic plots (Residuals vs Fitted, Q-Q, Scale-Location, Cook's D)
par(mfrow = c(2, 2))
plot(lm_af_main, which = 1:4)
par(mfrow = c(1, 1))
```

The residuals vs fitted plot shows a U-shaped pattern, suggesting a violation of the linearity assumption. Variance also appears non-constant across the fitted range. Observations 53, 78, and 85 are flagged as potentially influential by Cook's distance. The Q-Q plot is broadly acceptable, with minor deviations at the tails.

```{r af-multicollinearity}
# Variance Inflation Factors — check for multicollinearity among predictors
vif_af <- vif(lm_af_main)
cat("VIF values (AF model):\n")
print(round(vif_af, 3))
# VIF > 10 (or conservatively > 5) indicates a problematic level of collinearity
```

All GVIF^(1/(2·Df)) values are close to 1 (treatment: 1.022, age: 1.058, sex: 1.027), indicating no meaningful multicollinearity among predictors.

## · PR: Logistic Regression Diagnostics

Logistic regression assumptions differ from linear regression. We check for linearity, multicollinearity, complete separation, overall goodness of fit, and influential observations.

```{r pr-linearity}
# Box-Tidwell test: checks whether continuous predictors have a linear relationship
# with the log-odds of PR. Interaction term age:log(age) should be non-significant.
df_bt <- df |>
  filter(!is.na(age), !is.na(sex)) |>
  mutate(age_log_age = age * log(age))

glm_bt <- glm(PR ~ treatment + age + sex + age_log_age,
              data = df_bt, family = binomial)

bt_coef <- tidy(glm_bt) |> filter(term == "age_log_age")
cat(sprintf(
  "Box-Tidwell interaction (age × log(age)): beta = %.4f, p = %.4f\n",
  bt_coef$estimate, bt_coef$p.value
))
```

The Box-Tidwell test is borderline significant (β = 1.48, p = 0.046), suggesting a possible non-linear relationship between age and the log-odds of PR, though the small sample size warrants cautious interpretation.

```{r pr-separation}
# Check for complete or quasi-complete separation — a common issue with binary outcomes
# If any cell in the cross-tab is 0, separation may be present
cat("Cross-tabulation of treatment by PR:\n")
print(table(df$treatment, df$PR))

cat("\nCross-tabulation of sex by PR:\n")
print(table(df$sex, df$PR))
```

All cells are non-zero across both cross-tabulations (minimum cell count = 10), so there is no evidence of complete or quasi-complete separation.

```{r pr-goodness-of-fit}
# Hosmer-Lemeshow goodness-of-fit test
# H0: the model fits adequately; p > 0.05 is desired
hl_test <- hoslem.test(
  x = glm_pr_main$y,       # 直接从模型对象取，长度保证一致
  y = fitted(glm_pr_main),
  g = 10
)
cat(sprintf(
  "Hosmer-Lemeshow test: X² = %.4f, df = %d, p = %.4f\n",
  hl_test$statistic, hl_test$parameter, hl_test$p.value
))
```

The Hosmer-Lemeshow test is non-significant (χ² = 8.61, df = 8, p = 0.376), indicating adequate model fit.

```{r pr-multicollinearity}
# Variance Inflation Factors for the PR model
vif_pr <- vif(glm_pr_main)
cat("VIF values (PR model):\n")
print(round(vif_pr, 3))
```

All GVIF^(1/(2·Df)) values are close to 1 (treatment: 1.008, age: 1.071, sex: 1.059), indicating no meaningful multicollinearity.

```{r pr-influential}
# Cook's distance — identify influential observations
cooksd_pr <- cooks.distance(glm_pr_main)
influential_pr <- which(cooksd_pr > 4 / nrow(df))   # common rule of thumb: 4/n

cat(sprintf(
  "Number of potentially influential observations (Cook's D > 4/n): %d\n",
  length(influential_pr)
))

if (length(influential_pr) > 0) {
  cat("Participant IDs:", paste(df$ID[influential_pr], collapse = ", "), "\n")
}

# Visualise Cook's distances
tibble(
  obs     = seq_along(cooksd_pr),
  cooksd  = cooksd_pr,
  flagged = cooksd_pr > 4 / nrow(df)
) |>
  ggplot(aes(x = obs, y = cooksd, colour = flagged)) +
  geom_segment(aes(xend = obs, yend = 0), linewidth = 0.6) +
  geom_point(size = 2) +
  geom_hline(yintercept = 4 / nrow(df),
             linetype = "dashed", colour = "grey40") +
  scale_colour_manual(values = c(`FALSE` = "#0D7680", `TRUE` = "#E8A838"),
                      guide  = "none") +
  labs(
    title    = "Cook's Distance — PR Logistic Model",
    subtitle = "Dashed line = 4/n threshold; orange points warrant further inspection",
    x        = "Observation index",
    y        = "Cook's distance"
  ) +
  theme_ppt

save_ppt_fig("ppt_pr_cooks.png", width = 9, height = 5)
```
Seven observations exceed the 4/n threshold (participants 16, 40, 49, 57, 72, 82, 92), with participants 49 and 82 showing the largest Cook's distances. These should be examined for data entry errors, though removal is not recommended without substantive justification.
